\propertitle

### 2019-11-15 maxNt_v6.tex
update

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-11-15T19:22:27+0100
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal network\\ from subnetwork data {\large\color{myred}[draft]}%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

As mentioned in the Introduction (see references there), in the
neurosciences the maximum-entropy method has also been used as a way of
quantifying the \enquote{cooperativity} \citep[\eg][]{gersteinetal1985} or
\enquote{interaction}
\citep[\eg][]{martignonetal1995,schneidmanetal2006,shlensetal2006} or
\enquote{synchrony} \citep[\eg][we're only citing the first papers using
these terms]{bohteetal2000,amarietal2003} of neuronal
activity. In this section we discuss how our proposed
application bears on this kind of quantification. % For concreteness we use an
% example with specific statistics: first and second moments (equivalent to
% mean and correlations), versus first-to-fourth moments. But our
% discussion holds for any sets of statistics.

Let us translate this question into more precise terms first, because
\enquote{cooperativity}, \enquote{interaction}, and similar notions are
vague. A translation in terms of \emph{informational sufficiency} seems to
suit all these notions \parentext{for the technical definition and
  intuitive understanding of sufficiency, which goes back to
  \cite{fisher1922}, see \eg\ \cite[\sect~4.5]{bernardoetal1994};
  \cite[\chap~8, \sect~14.2]{jaynes1994_r2003}; \cite{cifarellietal1982};
  \cite{kullbacketal1951}. For technical results and connections with the
  notion of symmetry see \eg\
  \cite{darmois1935,neyman1935,koopman1936,pitman1936,halmosetal1949,bahadur1954,berk1972,lauritzen1974,cifarellietal1980,cifarellietal1981,diaconisetal1981,lauritzen1982_r1988,diaconis1992,furmanczyketal1998,fortinietal2000,nogalesetal2000,kallenberg2005,lauritzen2007,ayetal2015}}.
The idea is as follows. We want to assess our probabilities about the
frequencies of the activities of the sample, or about the activity of the
sample in a new time bin. For this assessment we should use all measured
data and statistics we have. But it can be the case that dropping part of
the data or statistics -- for example, the measured third- and higher-order
moments -- leaves our probabilities almost unchanged. This means that the dropped
statistics are \emph{informationally irrelevant} or almost so. The
remaining statistics -- for example, first and second moments -- are then
\emph{informationally
  sufficient}. % This informational quasi-sufficiency is interesting because
% it may give clues about network organization and dynamics.

We can thus quantify how much a subset of statistics is relevant with
respect to a larger set by keeping only the subset in the conditional of
our probabilities, and checking how much the latter change. When our
probabilities are built with the maximum-entropy method, this means keeping
only part of the constraints (and re-maximizing the entropy). For example,
keeping only the first two moments (means and correlations) out of the
first four moments. This procedure is actually equivalent to comparing the
probabilities of our hypotheses about sufficiency, conditional the
\emph{full} data, assuming their pre-data probabilities to be equal. We
show this equivalence below and perform the calculations for our first data
set.

% For example: does knowledge about
% third-order correlations change our state of uncertainty about network
% activity in an appreciable way, compared to knowledge about second-order
% correlations only? The method is used as follows to answer this question.
% We imagine to be ignorant of the measured frequency distribution for the
% network activity, and to know only up to second-order measured statistics,
% or up to third-order. Two maximum-entropy distributions are then built,
% constrained under the two sets of moments under comparison. Finally, the
% resulting distributions are compared with each other and with the measured
% frequency distribution. This recipe generalizes to any subsets of measured
% statistics, of course.

Before applying this idea to neuronal activity, however, it is important to
keep in mind that \emph{informational sufficiency is not preserved under
  sampling}: if a probability distribution has some sufficient statistics,
then its marginals \emph{cannot} have the same sufficient statistics, and
vice versa, except for trivial cases such as uniform probability
distributions. This generally also holds among different marginals: if a
statistics is sufficient for some marginal, then it will not be sufficient
for another marginal. This impossibility is known in statistical mechanics:
marginals of Gibbs states are not Gibbs states \citep{maesetal1999}.
Mathematically this impossibility comes from the Pitman-Koopman theorem
\citep{koopman1936,pitman1936,darmois1935} connecting sufficient statistics
and the exponential family, and translates into the general impossibility
of solving a system of $n$ independent equations in $m$ unknowns with more
equations than unknowns, $n>m$ \citep[\sect~3.1]{portamanaetal2015}.

This fact is important in the analysis of neural activity when it is based
on small samples from brain regions. If means and pairwise correlations
seem informationally sufficient for a particular sample from a brain area,
then they may well be \emph{insufficient} for the full network of neurons
constituting that area. Vice versa, if they are informationally sufficient
for the full network, then they may turn out \emph{not} to be sufficient
(or less so) for any recorded sample thereof. Hence, if we wonder about
\enquote{cooperativity} or \enquote{interaction} of a brain area, we cannot
reliably use maximum-entropy distributions \emph{for a sample only}. Yet
the neuroscientific literature is besprinkled with such sample-level
analyses
\citep[\eg][]{martignonetal1995,bohteetal2000,schneidmanetal2006,shlensetal2006,barreiroetal2010,ganmoretal2011,granotatedgietal2013}.

Questions about informational sufficiency are therefore correctly addressed
by applying the maximum-entropy method at the full-network level. If a
comparison with the measured frequencies of the sample activities is
desired, then the marginal distribution to the sample size should be used.

>>>>

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-11-15T21:36:39+0100
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal network\\ from subnetwork data {\large\color{myred}[draft]}%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

As mentioned in the Introduction (see references there), in the
neurosciences the maximum-entropy method has also been used as a way of
quantifying the \enquote{cooperativity} \citep[\eg][]{gersteinetal1985} or
\enquote{interaction}
\citep[\eg][]{martignonetal1995,schneidmanetal2006,shlensetal2006} or
\enquote{synchrony} \citep[\eg][-- we're only citing early papers using
these terms]{bohteetal2000,amarietal2003} of neuronal
activity. In this section we discuss how our proposed
application bears on this kind of quantification. % For concreteness we use an
% example with specific statistics: first and second moments (equivalent to
% mean and correlations), versus first-to-fourth moments. But our
% discussion holds for any sets of statistics.

\enquote{Cooperativity}, \enquote{interaction}, and similar terms are
vague, so we need to translate them into a more precise notion first. We
propose the notion of \emph{informational sufficiency}
\citep[\sect~4.5]{bernardoetal1994}[\chap~8 \amp\
\sect~14.2]{jaynes1994_r2003}{cifarellietal1982,kullbacketal1951}[the
notion goes back to][]{fisher1922} because it relates to those terms, is
intuitive, and connects directly to probability theory and maximum-entropy
distributions. Its idea is as follows. Our probabilities about the
frequencies of the activities of the sample, or about the activity of the
sample in a new time bin, are in principle conditional on all experimental
data and statistics we have. But it can be the case that discarding part of
the data or statistics -- for example, the measured third- and higher-order
moments -- leaves our probabilities almost unchanged. This means that the
discarded statistics are \emph{informationally irrelevant} or almost so.
The remaining statistics -- for example, first and second moments -- are
\emph{informationally sufficient}. \parentext{For more technical results
  and connections with the notion of symmetry see \eg\
  \cite{darmois1935,neyman1935,koopman1936,pitman1936,halmosetal1949,bahadur1954,berk1972,lauritzen1974,cifarellietal1980,cifarellietal1981,diaconisetal1981,lauritzen1982_r1988,diaconis1992,furmanczyketal1998,fortinietal2000,nogalesetal2000,kallenberg2005,lauritzen2007,ayetal2015}.}
% This informational quasi-sufficiency is interesting because
% it may give clues about network organization and dynamics.

We can thus quantify the informational relevance of a subset of statistics,
for example first and second moments (means and correlations), with respect
to a larger set, for example the first four moments, by comparing the
probabilities conditional on the subset and on the full set. For
probabilities built with the maximum-entropy method, this means comparing
those constrained on the subset and on the full set. This procedure is
actually equivalent to comparing the probabilities of the two hypotheses
about sufficiency, conditional the \emph{full} data, assuming their
pre-data probabilities to be equal. We show this equivalence below and
perform the calculations for our first data set.

Before applying this notion to neuronal activity, however, we must keep in
mind that \emph{informational sufficiency is not preserved under sampling}:
if a probability distribution has some sufficient statistics, then its
marginals, such as the distribution for a sample, \emph{cannot} have the
same sufficient statistics, and vice versa; except for trivial cases such
as uniform probability distributions. % This generally holds also among
% different marginals: if a statistics is sufficient for some marginal, then
% it will not be sufficient for another marginal.
This impossibility is known in statistical mechanics: if a system is
described by a Gibbs state, its subsystems cannot be perfectly described by
Gibbs states \citep[\eg][and references therein]{maesetal1999}.
Mathematically this impossibility comes from the Pitman-Koopman theorem
\citep{koopman1936,pitman1936,darmois1935} connecting sufficient statistics
and the exponential family of distributions, and translates into the
general impossibility of solving a system of $n$ independent equations in
$m$ unknowns with more equations than unknowns, $n>m$
\citep[\sect~3.1]{portamanaetal2015}.

This fact is important for our analysis. If, say, means and pairwise
correlations seem informationally sufficient for a particular sample from a
brain area, then they may well be \emph{insufficient} for the full network
of neurons constituting that area, and vice versa. If what interests us is
\enquote{cooperativity} or \enquote{interaction} of a brain area, we cannot
therefore reliably use maximum-entropy distributions \emph{for a sample
  only} thereof. Yet the neuroscientific literature is besprinkled with
such sample-level analyses
\citep[\eg][]{martignonetal1995,bohteetal2000,schneidmanetal2006,shlensetal2006,barreiroetal2010,ganmoretal2011,granotatedgietal2013}.
The approach presented here bypasses this problem instead, because the
maximum-entropy method is applied to obtain the distribution of the full
network, not of the sample alone.
